---
title: "Predicting Life Expectancy: How long people live? Why? Why not?"
subtitle: "STAT 420, Summer 2020"
author: "Venslaus Prakash Arokiaraj (netid: vpa2) & Pushpit Saxena(netid: pushpit2)"
date: 'July 18th, 2020'
output:
  html_document: 
    theme: readable
    highlight: tango
    toc: yes
  pdf_document: default
urlcolor: dodgerblue
---

<style>
body {
text-align: justify;
}
</style>

***
## Team Information

- Name: **team_bbg**

- Size: **`2`**

- Details:

| Member Name                | Member NetId |
|----------------------------|--------------| 
| Pushpit Saxena             | `pushpit2`   |
| Venslaus Prakash Arokiaraj | `vpa2`       |


# Please note that this is an interim report that we are submitting on coursera, as Prof. David has given us an extension till Sunday (08/09) to submit the final report (via. piazza). Please see [here](ExtensionConfirmation.pdf) for details. *(Final report will have more cleaner structure and some of the experimental models that you see in this will not be in final report. Please use that report for grading.)*  


## Introduction

This project will mainly focus on studying different factors that play statistically significant role in influencing **Life Expectancy**. Some of the factors we will be focussing on are economic factors, social factors, health services factors (like immunizzation levels), mortality rate and various other health related factors. We will be building different multiple linear regression models and will try to apply some of concepts that we have learned as part of this course (STAT 420) to analyze and find the appropriate models for predicting life expectancy.

 
# Dataset

## Background

Based on the description of the dataset on [kaggle](https://www.kaggle.com/kumarajarshi/life-expectancy-who), the Global Health Observatory(**GHO**) data repository under World Health Organization (**WHO**) keeps track of the health status as well as many other related factors for all countries The datasets are made available to public for the purpose of health data analysis. This datset was collected from WHO and United Nations website and then the individual data files have been combined into a single data set (read more [here](https://www.kaggle.com/kumarajarshi/life-expectancy-who#Description))

## Description

The dataset we will be using for this project is Life Expectancy data that can be found at [Life Expectancy (WHO)](https://www.kaggle.com/kumarajarshi/life-expectancy-who?select=Life+Expectancy+Data.csv). The dataset has 22 variables and  2939 observations which needs some cleanup. (*Note: we have also provided the dataset as part of the .zip [[`lifeExpectancyData`](lifeExpectancyData.csv)] that we have uploaded for this proposal*).

Following are some of the important variables used in this dataset:

- **`Country`** (String): Country of observation

- **`Year`** (Integer): Year of observation

- **`Status`** (String): Whether the country of observation is developed or developing.

- **`Life expectancy`** (Decimal): Life expectancy in age

- **`Adult Mortality`** (Integer): Adult Mortality Rates of both sexes (probability of dying between 15 and 60 years per 1000 population) 

- **`Infant deaths`** (Integer): Number of Infant Deaths per 1000 population

- **`Alcohol`** (Decimal): Alcohol, recorded per capita (15+) consumption (in litres of pure alcohol) 

- **`Percentage Expenditure`** (Decimal): Expenditure on health as a percentage of Gross Domestic Product per capita(%)

- **`Hepatitis B`** (Int): Hepatitis B (HepB) immunization coverage among 1-year-olds (%) 

- **`Measles`** (Int): Measles - number of reported cases per 1000 population 

- **`BMI`** (Decimal): Average Body Mass Index of entire population

- **`Under-five deaths`** (Int): Number of under-five deaths per 1000 population

- **`Polio`** (Int): Polio (Pol3) immunization coverage among 1-year-olds (%)

- **`Total expenditure`** (Decimal): General government expenditure on health as a percentage of total government expenditure (%)

- **`Diphtheria`** (Int): Diphtheria tetanus toxoid and pertussis (DTP3) immunization coverage among 1-year-olds (%)

- **`HIV/AIDS`** (Decimal): Deaths per 1 000 live births HIV/AIDS (0-4 years)

- **`GDP`** (Decimal): Gross Domestic Product per capita (in USD)

- **`Population`** (Int): Population of the country

# Data Source

We have attached the data file in the **`.zip`** [[`lifeExpectancyData`](lifeExpectancyData.csv)] .

If needed, this dataset can also be downloaded from [kaggle](https://www.kaggle.com/kumarajarshi/life-expectancy-who)

# Our Motivation

- From the standpoint of our research and learning, one of the primary reason for us to pick this dataset, is the size as well as variety of predictors that are available. We believe that this dataset is perfect for us to practice and implement majority of the techniques we have learned as part of the course and get a hands-on experience on a real life dataset.

- From larger perspective of data exploration and using data science to address real world issues, this dataset gives us an opportunity to try and answer some of the most important questions the human race is facing, like various factors affecting the longevity of life. As we briefly mentioned in our project description, we are interested in determining different factors which contributes to lower the value of life expectancy. Particularly, in this dataset as the observations are based on different countries, if we are able to find a good model than we can answer questions like what a country needs to focus on in order to have better life expectancy.

# Data (Evidence)

- Reading the data file:

```{r}
life_expectancy_data <- read.csv("lifeExpectancyData.csv")
```

- Data snippet

```{r}
library(tibble)
as_tibble(life_expectancy_data)
```

- First few values of `Life Expectancy` (response):

```{r}
life_expectancy_data[1:10, c("Life.expectancy")]
```


# Methods and Results

## Data Cleaning:

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyr)
library(tidyverse)
library(countrycode)
library(lemon)
library(skimr)
library(naniar)
library(MASS)
knit_print.data.frame <- lemon_print
```

**Loading the Data:**

```{r message=FALSE, warning=FALSE}
raw_data <- read.csv("LifeExpectancyData.csv")

# Added Continent
raw_data$Continent <- countrycode(sourcevar = raw_data[, "Country"],
                              origin = "country.name",
                              destination = "continent")
# Added Region
raw_data$region <- countrycode(sourcevar = raw_data[, "Country"],
                              origin = "country.name",
                              destination = "region")

```


```{r echo=FALSE}
raw_data$Country <- as.factor(raw_data$Country)
raw_data$Status <- as.factor(raw_data$Status)
raw_data$Continent <- as.factor(raw_data$Continent)
raw_data$Year <- as.factor(raw_data$Year)
raw_data$region <- as.factor(raw_data$region)


calc_rmse<- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2))
}

```

**Changing the names of the fields to follow a more consistent pattern(snake-case):**

```{r}
col_names <- tolower(trimws(str_replace_all(colnames(raw_data), "\\.+", "_")))
# col_names <- tolower(str_replace_all(colnames(raw_data), "\\s+", ""))
colnames(raw_data) <- col_names
```

**Snippet of the raw dataset:**

```{r}
tibble::as.tibble(raw_data)
```

**Summary of `numeric` fields:**

```{r echo = FALSE, warning=FALSE, message=FALSE}
# summary_vector <- summary(raw_data %>% select_if(negate(is.factor)))
# data.frame(unclass(summary_vector), check.names = FALSE, stringsAsFactors = FALSE)
knitr::kable(t(do.call(cbind, lapply(raw_data %>% select_if(negate(is.factor)), summary))),
             format.args = list(scientific = FALSE),
             digits = 2)
```

We can see that only `10` observations have missing values for the response field `life_expectancy`, so we drop those `10` observations as dropping them will not make much difference to the models that we will try. 
```{r}
mod_data_df <- raw_data[!is.na(raw_data$life_expectancy),]
nrow(mod_data_df)
```

There are still `r nrow(mod_data_df[!complete.cases(mod_data_df),])` observations with some missing values. 
We will use the mean of the value for a given `country` to impute some of these values:
```{r message=FALSE, warning=FALSE}
new_df <- mod_data_df %>% group_by(country) %>% mutate_if(is.numeric,
                                              function(x) ifelse(is.na(x), mean(x, na.rm = TRUE), x))
nrow(new_df[!complete.cases(new_df),])
```

Still there are some observations with missing values. Next we will use the mean of the values for a given `region` in a particular `year` to impute some of these missing values:

```{r message=FALSE, warning=FALSE}
cleaned_df <- as.data.frame(new_df %>% group_by(region, year) %>% mutate_if(is.numeric,
                                              function(x) ifelse(is.na(x), mean(x, na.rm = TRUE), x)) %>% ungroup)
cleaned_df$region <- as.factor(cleaned_df$region)
cleaned_df$year <- as.factor(cleaned_df$year)
nrow(cleaned_df[!complete.cases(cleaned_df),])
```

Finally, we have imputed all the values and our final dataset has ``r nrow(cleaned_df)`` observations

## Data exploration and visualization

- Basic statistics (by region)

```{r message=FALSE, warning=FALSE, echo = FALSE}
dd <- as.data.frame(
raw_data %>%
  group_by(region) %>%
  summarize(count = n(),
            mean_life_expectancy = mean(life_expectancy, na.rm = TRUE),
            mean_infant_deaths   = mean(infant_deaths, na.rm = TRUE),
            mean_adult_deaths    = mean(adult_mortality, na.rm = TRUE)
            ))

colnames(dd) <- c("Region", "#Records", "Avg. Life Expectancy", "Avg. Infant Deaths", "Avg. Adult Deaths")
knitr::kable(dd)
```

```{r message=FALSE, warning=FALSE, echo = FALSE}
dd <- as.data.frame(
raw_data %>%
  group_by(continent) %>%
  summarize(count = n(),
            mean_life_expectancy = median(life_expectancy, na.rm = TRUE),
            mean_infant_deaths   = mean(infant_deaths, na.rm = TRUE),
            mean_adult_deaths    = mean(adult_mortality, na.rm = TRUE)
            ))
dd
```


```{r}
library(ggplot2)
ggplot(data = dd, aes(x = continent, y = mean_life_expectancy)) + geom_bar(stat = "identity")

ggplot(data = raw_data, aes(x = continent, y = life_expectancy)) + geom_boxplot(aes(fill = continent))

```

```{r fig.width=10, fig.height=10, fig.align="center"}
library(reshape2)
library(ggcorrplot)
cor_mat <- round(cor(na.omit(raw_data[, !(names(raw_data) %in% c("year",
"country", "continent", "status", "region"))])), 1)
cor_mat
ggcorrplot(cor_mat, lab = TRUE)
# ggplot(data = melted_cormat, aes(x = Var1, y = Var2, fill = value)) + geom_tile()
```

***

## Model Building

Splitting the data in training and test set (`90%` training, `10%` hold out test set):

```{r}
set.seed(19851115)
le_trn_data_idx  <- sample(nrow(cleaned_df), size = trunc(0.90 * nrow(cleaned_df)))
le_trn_data <- cleaned_df[le_trn_data_idx, ]
le_tst_data <- cleaned_df[-le_trn_data_idx, ]
```

Ignoring all the categorical variables for now (except `status`, we have fitted models using some of these categorical variables but couldn't get better results, code can be seen in Appendix)

```{r}
non_cat_predictor_df <- subset(le_trn_data, select = -c(year, country, continent, region))
```

We started with fitting a full Additive model (with all the numerical predictor and `status`). This will provide us with a good baseline model to do simple as well as more nuanced feature selections later

```{r}
full_additve_model <- lm(life_expectancy ~ ., data = non_cat_predictor_df)
```

- Summary of the full additive model

```{r}
summary(full_additve_model)
```

- Diagnostics plots for full additive model

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.align="center"}
par(mfrow = c(2, 2))
plot(full_additve_model, col="darkorange")
```

- Test $\mathbf{\text{RMSE}} = `r calc_rmse(le_tst_data$life_expectancy, predict(full_additve_model, newdata = le_tst_data))`$
- $R^2 = `r summary(full_additve_model)$r.squared`$
- We can see that this model show that there are some non-significant predictors in the full model, for e.g. for `alcohol`, if we use t-test for significance:
  - Null: $H_0:$ $\beta_{alcohol} = 0$
  - Alternative: $H_0:$ $\beta_{alcohol} \neq 0$
  - Test Statistics: $`r summary(full_additve_model)$coefficient["alcohol", "t value"]`$
  - P-value: $`r summary(full_additve_model)$coefficient["alcohol", "Pr(>|t|)"]`$
  - Decision: **Fail** to reject null
  - Conclusion: `alcohol` does not have significant linear relationship with `life_expectancy`
- Also, we can see from the diagnostic plots that both equal variance assumption and normality assumption are suspect.


So we started with simple (not recommended) method of removing some of the least significant predictors. Also, there seems to be high collinearity between `infant_deaths` and `under_5_deaths` (check `vif` below and `correlation plot` shown earlier).

```{r}
car::vif(full_additve_model)
```

So we removed some of the least significant predictor and kept `infant_deaths`

```{r}
sig_additive_model <- lm(life_expectancy ~  adult_mortality +
                         infant_deaths + bmi + diphtheria + hiv_aids + gdp +
                         income_composition_of_resources * status  + schooling,
                       data = non_cat_predictor_df)
summary(sig_additive_model)
```

- Diagnostic plot:

```{r fig.align="center", echo=FALSE}
par(mfrow = c(2, 2))
plot(sig_additive_model, col="darkorange", )
```

- Test $\mathbf{\text{RMSE}} = `r calc_rmse(le_tst_data$life_expectancy, predict(sig_additive_model, newdata = le_tst_data))`$
- $R^2 = `r summary(sig_additive_model)$r.squared`$

```{r fig.align="center"}
sig_interative_model <- lm(life_expectancy ~  (adult_mortality +
                         under_five_deaths + bmi  + diphtheria + hiv_aids + gdp  +
                         income_composition_of_resources + schooling) ^ 2 ,
                       data = non_cat_predictor_df)
summary(sig_interative_model)
par(mfrow = c(2, 2))
plot(sig_interative_model, col="darkorange", )
calc_rmse(le_tst_data$life_expectancy,
          predict(sig_interative_model, newdata = le_tst_data))
```


```{r}
# aic_model_backward <- step(sig_linear_model, direction = "backward")
# summary(aic_model_backward)
# par(mfrow = c(2, 2))
# plot(aic_model_backward, col="darkorange", )
# calc_rmse(le_tst_data$life_expectancy,
#           predict(aic_model_backward, newdata = le_tst_data))
```


```{r}
# aic_model_both <- step(sig_linear_model, direction = "both")
# summary(aic_model_both)
# par(mfrow = c(2, 2))
# plot(aic_model_both, col="darkorange", )
# calc_rmse(le_tst_data$life_expectancy,
#           predict(aic_model_both, newdata = le_tst_data))
```


```{r}
# bic_model_both <- step(sig_linear_model, k = log(length(resid(sig_linear_model))), direction = "both")
# summary(bic_model_both)
# par(mfrow = c(2, 2))
# plot(bic_model_both, col="darkorange", )
# calc_rmse(le_tst_data$life_expectancy,
#           predict(bic_model_both, newdata = le_tst_data))
```


```{r}
# bc <- boxcox(sig_linear_model, lambda = seq(-3, 3))
# best.lam <- bc$x[which(bc$y==max(bc$y))]
```


```{r}
# sig_linear_model_sq <- lm(life_expectancy ^ 2 ~ adult_mortality + alcohol + percentage_expenditure +
#                          under_five_deaths + bmi + diphtheria + hiv_aids + gdp +
#                          income_composition_of_resources  + schooling, data = non_cat_predictor_df)
# summary(sig_linear_model_sq)
# par(mfrow = c(2, 2))
# plot(sig_linear_model_sq, col="darkorange", )
# shapiro.test(resid(sig_linear_model_sq))
# calc_rmse(le_tst_data$life_expectancy,
#           predict(sig_linear_model_sq, newdata = le_tst_data))
```

```{r}
# aic_model <- step(sig_linear_model, direction = "both", test = "F")
# par(mfrow = c(2, 2))
# plot(aic_model)
```

```{r}
lm_model <- lm(life_expectancy ~ income_composition_of_resources + adult_mortality + 
                  bmi +  status + under_five_deaths,
               data = non_cat_predictor_df)
par(mfrow = c(2, 2))
plot(lm_model)
calc_rmse(le_tst_data$life_expectancy,
          predict(lm_model, newdata = le_tst_data))
```

```{r}
bc <- boxcox(lm_model, lambda = seq(-3, 3))
best.lam <- bc$x[which(bc$y==max(bc$y))]
```

```{r}
lm_model_cube <- lm(life_expectancy ^ 3 ~ income_composition_of_resources + adult_mortality + 
                  bmi +  status + under_five_deaths,
               data = non_cat_predictor_df)
summary(lm_model_cube)
par(mfrow = c(2, 2))
plot(lm_model_cube)
calc_rmse(le_tst_data$life_expectancy,
          predict(lm_model_cube, newdata = le_tst_data))
```



```{r}
model1 <- lm(life_expectancy ~ schooling + bmi + alcohol + gdp + hiv_aids + diphtheria + status, data = non_cat_predictor_df)
summary(model1)
par(mfrow = c(2, 2))
plot(model1)
calc_rmse(le_tst_data$life_expectancy,
          predict(model1, newdata = le_tst_data))
```

```{r}
model2<-step(model1, data=non_cat_predictor_df, direction = "both", test="F")
summary(model2)
par(mfrow = c(2, 2))
plot(model2)
calc_rmse(le_tst_data$life_expectancy,
          predict(model2, newdata = le_tst_data))
```


```{r}
model1 <- lm(life_expectancy ~ schooling + bmi + gdp * status + hiv_aids + diphtheria, data = non_cat_predictor_df)
summary(model1)
par(mfrow = c(2, 2))
plot(model1)
calc_rmse(le_tst_data$life_expectancy,
          predict(model1, newdata = le_tst_data))
```
<!-- ```{r} -->

```{r fig.width=10, fig.height=5, }
bp_df <- subset(non_cat_predictor_df, 
                select = c(schooling,bmi,hiv_aids,diphtheria))
boxplot(bp_df)

boxplot(non_cat_predictor_df$gdp)
```


```{r}
model1 <- lm(life_expectancy ~ (schooling + bmi  + gdp + hiv_aids + diphtheria) * status, data = le_trn_data)
summary(model1)
par(mfrow = c(2, 2))
plot(model1)
```
```{r}
model2 <- step(model1, direction = "backward")
summary(model2)
par(mfrow = c(2, 2))
plot(model2)
```

<!-- ``` -->

```{r}
library(leaps)
regs <- regsubsets(life_expectancy ~ ., data = non_cat_predictor_df, nbest=10)
plot(regs, 
     scale="adjr", 
     main="All possible regression: ranked by Adjusted R-squared")
```

```{r}
model1 <- lm(life_expectancy ~ adult_mortality + bmi + hiv_aids +
               income_composition_of_resources + schooling, data = le_trn_data)
summary(model1)
par(mfrow = c(2, 2))
plot(model1)
```

```{r}
boxplot(le_trn_data$life_expectancy, ylab = "Life Expectancy (Age)")
outliers_out <- boxplot(le_trn_data$life_expectancy, plot = F)$out # untuk mendaptkan outlier
life_clean <- le_trn_data[-which(le_trn_data$life_expectancy %in% outliers_out), ]
nrow(life_clean)
boxplot(life_clean$life_expectancy, ylab = "Life Expectancy (Age)") #

model1 <- lm(life_expectancy ~ adult_mortality + bmi + hiv_aids +
               income_composition_of_resources + schooling, data = life_clean)
summary(model1)
par(mfrow = c(2, 2))
plot(model1)

```

```{r}
sig_linear_model <- lm(life_expectancy ~  adult_mortality +
                         under_five_deaths + bmi + diphtheria + hiv_aids + gdp +
                         income_composition_of_resources + schooling + status,
                       data = life_clean)
summary(sig_linear_model)
par(mfrow = c(2, 2))
plot(sig_linear_model, col="darkorange", )
```

```{r}
full_model <- lm(life_expectancy ~ ., data = subset(life_clean, select = -c(year, country, continent, region)))
```

```{r fig.width=10, fig.height=10, fig.align="center"}
aic_model <- step(full_model, direction = "both")
summary(aic_model)
par(mfrow = c(2, 2))
plot(aic_model, col="darkorange", )
```

```{r fig.width=10, fig.height=10, fig.align="center"}
bic_model <- step(full_model, direction = "backward")
summary(bic_model)
par(mfrow = c(2, 2))
plot(bic_model, col="darkorange", )
```

```{r fig.width=10, fig.height=10, fig.align="center"}
clean_bic <- lm(life_expectancy ~ status + adult_mortality + infant_deaths + 
     + bmi + under_five_deaths +
    polio + diphtheria + hiv_aids + income_composition_of_resources + schooling,
    data = life_clean)
summary(clean_bic)
par(mfrow = c(2, 2))
plot(clean_bic, col="darkorange", )
```


```{r}
bc <- boxcox(clean_bic, lambda = seq(-3, 3))
best.lam <- bc$x[which(bc$y==max(bc$y))]
```


```{r}
life_clean <- life_clean[which(life_clean$life_expectancy > 50), ]
nrow(life_clean)
```

```{r fig.align="center"}
clean_bic <- lm(life_expectancy ~ status + adult_mortality + infant_deaths + 
     + bmi + under_five_deaths +
    polio + diphtheria + hiv_aids + income_composition_of_resources + schooling,
    data = life_clean)
summary(clean_bic)
par(mfrow = c(2, 2))
plot(clean_bic, col="darkorange", )
```


```{r}
life_clean1 <- subset(life_clean, select = -c(year, country, continent, region))
clean_full <- lm(formula = life_expectancy ~ ., data = life_clean1)
clean_none <- lm(formula = life_expectancy ~ 1, data = life_clean1)
```

```{r fig.align="center"}
clean_backward <- step(clean_full, direction = "backward")
summary(clean_backward)
par(mfrow = c(2, 2))
plot(clean_backward, col="darkorange")
```


```{r}
small_cleaner_backward <- lm(life_expectancy ~ status + adult_mortality + infant_deaths + bmi + under_five_deaths + 
     total_expenditure + diphtheria + hiv_aids + gdp +
    thinness_1_19_years + income_composition_of_resources + schooling,
    data = life_clean1)

summary(small_cleaner_backward)
par(mfrow = c(2, 2))
plot(small_cleaner_backward, col="darkorange")

calc_rmse(le_tst_data$life_expectancy, predict(small_cleaner_backward, newdata = le_tst_data))
```

```{r fig.width=10, fig.height=10, fig.align="center"}
# pairs(subset(life_clean1, select = c(life_expectancy, adult_mortality, infant_deaths, bmi,
#                                      under_five_deaths, total_expenditure, diphtheria, hiv_aids,
#                                      gdp, thinness_1_19_years, income_composition_of_resources,
#                                      schooling)), col = "deepskyblue")
```


```{r fig.width=10, fig.height=10}
small_cleaner_backward_log <- lm(life_expectancy ~ 
                                   status + log1p(adult_mortality) + log1p(bmi) +
                                   log1p(infant_deaths) + total_expenditure + diphtheria + hiv_aids +
                                   log1p(gdp) + log1p(thinness_1_19_years) + income_composition_of_resources + schooling,
    data = life_clean1)

summary(small_cleaner_backward_log)
par(mfrow = c(2, 2))
plot(small_cleaner_backward_log, col="darkorange")

calc_rmse(le_tst_data$life_expectancy, predict(small_cleaner_backward_log, newdata = le_tst_data))
library(lmtest)
bptest(small_cleaner_backward_log)
```

```{r}
small_cleaner_backward_log_log <- lm(log1p(life_expectancy) ~ 
                                   status + log1p(adult_mortality) + log1p(bmi) +
                                   log1p(under_five_deaths) + total_expenditure + diphtheria + hiv_aids +
                                   log1p(gdp) + log1p(thinness_1_19_years) + income_composition_of_resources + schooling,
    data = life_clean1)

summary(small_cleaner_backward_log_log)
par(mfrow = c(2, 2))
plot(small_cleaner_backward_log_log, col="darkorange")

calc_rmse(le_tst_data$life_expectancy, predict(small_cleaner_backward_log_log, newdata = le_tst_data))



```

```{r}
bc <- boxcox(small_cleaner_backward_log, lambda = seq(-3, 3))
best.lam <- bc$x[which(bc$y==max(bc$y))]
best.lam
small_cleaner_backward_box_cox <- lm(life_expectancy ^ 2 ~ 
                                   status + log1p(adult_mortality) + log1p(bmi) +
                                   log1p(under_five_deaths) + total_expenditure + diphtheria + hiv_aids +
                                   log1p(gdp) + log1p(thinness_1_19_years) + income_composition_of_resources + schooling,
    data = life_clean1)
library(pracma)
summary(small_cleaner_backward_box_cox)
par(mfrow = c(2, 2))
plot(small_cleaner_backward_box_cox, col="darkorange")
pred <- predict(small_cleaner_backward_box_cox, newdata = le_tst_data)
sum(pred < 0)
calc_rmse(le_tst_data$life_expectancy, sqrt(predict(small_cleaner_backward_box_cox, newdata = le_tst_data)))

```
```{r}
plot(life_expectancy ~ I(log1p(under_five_deaths)), data = life_clean1)
```

```{r fig.width=10}
small_cleaner_backward_log_poly <- lm(life_expectancy ~ 
                                   status + log1p(adult_mortality)  + I(bmi ^ 2) * status + 
                                   log1p(under_five_deaths) + log1p(total_expenditure) + 
                                     diphtheria + I(diphtheria ^ 2) + hiv_aids + I(hiv_aids ^ 2) +
                                   log1p(gdp)  + log1p(thinness_1_19_years) + income_composition_of_resources + I(income_composition_of_resources ^ 2) + I(income_composition_of_resources ^ 3) +
                                   schooling + I(schooling ^ 2) + I(schooling ^ 3),
    data = life_clean1)

summary(small_cleaner_backward_log_poly)
par(mfrow = c(2, 2))
plot(small_cleaner_backward_log_poly, col="darkorange")

calc_rmse(le_tst_data$life_expectancy, predict(small_cleaner_backward_log_poly, newdata = le_tst_data))
bptest(small_cleaner_backward_log_poly)
```


```{r}
small_cleaner_backward_log_poly <- lm(life_expectancy  ~ 
                                   status + log1p(adult_mortality)  + I(bmi ^ 2) * status + 
                                   log1p(infant_deaths) + I(infant_deaths ^ 2) + log1p(total_expenditure) + I(gdp ^ 2) +
                                     diphtheria + I(diphtheria ^ 2) + hiv_aids + I(hiv_aids ^ 2) +
                                   log1p(gdp)  + log1p(thinness_1_19_years) + income_composition_of_resources + I(income_composition_of_resources ^ 2) +
                                   schooling + I(schooling ^ 2),
    data = life_clean1)

summary(small_cleaner_backward_log_poly)
par(mfrow = c(2, 2))
plot(small_cleaner_backward_log_poly, col="darkorange")

calc_rmse(le_tst_data$life_expectancy, predict(small_cleaner_backward_log_poly, newdata = le_tst_data))

hist(life_clean1$life_expectancy)
bptest(small_cleaner_backward_log_poly)
```

```{r fig.width=10}
small_cleaner_backward_log_poly <- lm(life_expectancy ~ 
                                   (status + log1p(adult_mortality)  + I(bmi ^ 2) * status + 
                                   log1p(under_five_deaths) + log1p(total_expenditure) + 
                                     diphtheria + I(diphtheria ^ 2) + hiv_aids + I(hiv_aids ^ 2) +
                                   log1p(gdp)  + log1p(thinness_1_19_years) + income_composition_of_resources + I(income_composition_of_resources ^ 2) + I(income_composition_of_resources ^ 3) +
                                   schooling + I(schooling ^ 2) + I(schooling ^ 3)) ^ 2,
    data = life_clean1)

# summary(small_cleaner_backward_log_poly)
par(mfrow = c(2, 2))
plot(small_cleaner_backward_log_poly, col="dodgerblue")

calc_rmse(le_tst_data$life_expectancy, predict(small_cleaner_backward_log_poly, newdata = le_tst_data))
bptest(small_cleaner_backward_log_poly)

```

```{r}
life_clean_glm <- life_clean1
life_clean_glm$life_expectancy_class <- ifelse(life_clean_glm$life_expectancy > 65, "hi", "low")
life_clean_glm$life_expectancy_class <- as.factor(life_clean_glm$life_expectancy)
le_tst_data$life_expectancy_class <- ifelse(le_tst_data$life_expectancy > 65, "hi", "low")
le_tst_data$life_expectancy_class <- as.factor(le_tst_data$life_expectancy_class)

small_cleaner_backward_log_poly_logit <- glm(life_expectancy_class ~ 
                                   status + log1p(adult_mortality)  + I(bmi ^ 2) * status + 
                                   log1p(under_five_deaths) + log1p(total_expenditure) + 
                                     diphtheria + I(diphtheria ^ 2) + hiv_aids + I(hiv_aids ^ 2) +
                                   log1p(gdp)  + log1p(thinness_1_19_years) + income_composition_of_resources + I(income_composition_of_resources ^ 2) +
                                   schooling + I(schooling ^ 2),
    data = life_clean_glm, family = "binomial")

summary(small_cleaner_backward_log_poly_logit)
par(mfrow = c(2, 2))
plot(small_cleaner_backward_log_poly_logit, col="darkorange")

```


# Appendix

```{r}
fit <- lm(life_expectancy ~ ., data = subset(le_trn_data, select = -c(year, continent, status, country)))
summary(fit)
par(mfrow = c(2, 2))
plot(fit)
```

